Binning Strategies and Related Loss for Large Data

### 2015-04-15 Paper.Rnw
Move a paragraph and color new changes

Using a small number of bins removes much of the relevant information in the data, while an extremely large number of bins may not reduce the data size or complexity to a sufficient degree. Figure~\ref{binning} gives an overview of a data set and binned representations using different numbers of bins, demonstrating the loss of information with increasing bin size.

@@

Loss of information occurs during the binning and rendering process. For the remainder of the paper we will assume that we are using colors in binned scatterplots to represent frequencies.  We distinguish two sources of loss in the construction of a binned scatterplot:
\begin{itemize}
\item {\it Spatial Loss}, $L^\text{S}$, occurs when points $(x_i, y_i)$ for observations $i \in \{1,\dots,n\}$ in the data set are reduced to a set of tiles centered at $(x_\ell^\ast, y_\ell^\ast)$ for bins $\ell \in \{1,\dots,\mathscr{L}\}$. By displaying frequency information using tiles that are visually centered at the bin centers causes a loss of information about the exact location of the points.

\item {\it Frequency Loss}, $L^\text{F}$, occurs when bin counts, $c_\ell \in \{1,\dots,\mathscr{L}\}$ are not rendered precisely. A discrete color scale may be used to aid the ability to read a binned scatterplot, however this is effectively a second stage of binning where shade displays binned frequencies, $b_C(c_\ell) \in \{1,\dots,\mathscr{L}\}$, thus losing precision in the display of the true frequencies. This is analogous to the fundimental problem for overplotted standard scatterplots where frequency is masked by other points. 
\end{itemize}

@@

\paragraph{Traditional Scatterplots and Frequency Information: (old writeup - to be reworked or removed)}

Using a minimal bin width, a binned scatterplot is comparable to a standard scatter plot, with bins shaded in a binary manner, as each unique observed value is located in a different bin. Alpha blending as used in Figure~\ref{fig:scatter-alpha}c extends the binary shading of a standard scatterplot to an implicit shading according to frequency. The shading is implicit because the range of frequency information is not scaled to the range of shading values, so that maximum color saturation is reached well before the maximum frequency, truncating the perceivable frequency information. 
By explicitly shading bins according to frequency, more information is preserved than in a traditional scatter plot, as the frequency domain provides visual weight to tiles which represent more points. This generalization allows us to describe the plots in Figure~\ref{fig:scatter-alpha}d and e under the same framework as plots a-c.

By additionally increasing the bin width, we  provide increasingly higher-level summaries of the data by  smoothing over local structures. For small bin widths these structures are likely to be  {\it noise} inherent in any real data set, while removing structures associated with  larger bin widths likely masks real signal in the data.

@@

Large Data sets of continuous variables are very difficult to visualize in raw form, due to over-plotting of points. Binning allows for the visualization and manipulation of large data sets, and easily translates into binned scatterplots which are more appropriate for the human visual system. Reducing the data for binned scatterplots has distinct computational and visual advantages, however the aggregation comes at the cost of losing precision in the spatial and frequency information.

We have presented two algorithms for spatially binning data points; the standard algorithm, which has rounding artifacts but generally lower loss, and the random algorithm, which is a fast version of linear binning and is somewhat more computationally efficient. The standard algorithm is superior in terms of net spatial loss, but requires more care in selecting binning parameters than random binning to avoid generating problematic visual features. In setting binning parameters for each spatial dimension, binwidth should be an integer multiple of the data resolution and the origin should be offset half the data resolution below the lowest observed value.

We also have explored the loss of frequency information for cases where frequencies are also binned to create a discrete color scale for tile shading. It is recommended to use less than ten seperate shades due to bounds of perceptual discrimination but at least four shades to avoid the steep frequency losses seen with fewer frequency bins. Using quantile frequency binning and log transforming the bin counts prior to standard frequency binning are shown to be reasonable methods -- with slight different interpretability -- for handling situations with heavily skewed bin count distributions.

@@

\begin{appendix}
\section{Mathematical Considerations}

>>>>

\ktm{Using a minimal bin width, a binned scatterplot is comparable to a standard scatter plot, with bins shaded in a binary manner, as each unique observed value is located in a different bin. Alpha blending as used in Figure~\ref{fig:scatter-alpha}c extends the binary shading of a standard scatterplot to an implicit shading according to frequency. The shading is implicit because the range of frequency information is not scaled to the range of shading values, so that maximum color saturation is reached well before the maximum frequency, truncating the perceivable frequency information. By explicitly shading bins according to frequency, more information is preserved than in a traditional scatter plot, as the frequency domain provides visual weight to tiles which represent more points. This generalization allows us to describe the plots in Figure~\ref{fig:scatter-alpha}d and e under the same framework as plots a-c. }

\ktm{By additionally increasing the bin width, we provide increasingly higher-level summaries of the data by smoothing over local structures. Using a small number of large bins may mask the real signal in the data, while an extremely large number of small bins may not sufficiently smooth over {\it noise} inherent in any real data set. Figure~\ref{binning} gives an overview of a data set and binned representations using different numbers of bins, demonstrating the loss of information with increasing bin size. }

@@

Loss of information occurs during the binning and rendering process. For the remainder of the paper we will assume that we are using colors in binned scatterplots to represent frequencies.  We distinguish two sources of loss in the construction of a binned scatterplot:

\ktm{
\begin{itemize}
\item {\it Spatial Loss}, $L^\text{S}$, occurs when points $(x_i, y_i)$ for observations $i \in \{1,\dots,n\}$ in the data set are reduced to a set of tiles centered at $(x_\ell^\ast, y_\ell^\ast)$ for bins $\ell \in \{1,\dots,\mathscr{L}\}$. By displaying frequency information using tiles that are visually centered at the bin centers causes a loss of information about the exact location of the points.

\item {\it Frequency Loss}, $L^\text{F}$, occurs when bin counts, $c_\ell \in \{1,\dots,\mathscr{L}\}$ are not rendered precisely. A discrete color scale may be used to aid the ability to read a binned scatterplot, however this is effectively a second stage of binning where shade displays binned frequencies, $b_C(c_\ell) \in \{1,\dots,\mathscr{L}\}$, thus losing precision in the display of the true frequencies. This is analogous to the fundimental problem for overplotted standard scatterplots where frequency is masked by other points. 
\end{itemize}
}

@@

@@

\ktm{Large Data sets of continuous variables are very difficult to visualize in raw form, due to over-plotting of points. Binning allows for the visualization and manipulation of large data sets, and easily translates into binned scatterplots which are more appropriate for the human visual system. Reducing the data for binned scatterplots has distinct computational and visual advantages, however the aggregation comes at the cost of losing precision in the spatial and frequency information. }

\ktm{We have presented two algorithms for spatially binning data points; the standard algorithm, which has rounding artifacts but generally lower loss, and the random algorithm, which is a fast version of linear binning and is somewhat more computationally efficient. The standard algorithm is superior in terms of net spatial loss, but requires more care in selecting binning parameters than random binning to avoid generating problematic visual features. In setting binning parameters for each spatial dimension, binwidth should be an integer multiple of the data resolution and the origin should be offset half the data resolution below the lowest observed value. }

\ktm{We also have explored the loss of frequency information for cases where frequencies are also binned to create a discrete color scale for tile shading. It is recommended to use less than ten seperate shades due to bounds of perceptual discrimination but at least four shades to avoid the steep frequency losses seen with fewer frequency bins. Using quantile frequency binning and log transforming the bin counts prior to standard frequency binning are shown to be reasonable methods -- with slight different interpretability -- for handling situations with heavily skewed bin count distributions.  }

@@

\begin{appendix}
\section{Appendix}
